Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-19 paper/main.tex
update nb, more edits

\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Black-box combinatorial optimization problems arise in practice when the
  objective function is evaluated by means of a simulator or a real-world
  experiment. In such cases, classical techniques such as mixed-integer
  programming and local search cannot be applied. Moreover, each fitness
  evaluation is often expensive in terms of time or resources, thus only a
  limited number of evaluations is possible, typically several orders of
  magnitude smaller than in white-box optimization problems. In the continuous
  case, Bayesian optimization, in particular using Gaussian processes, has
  proven very effective under these conditions. Much less research is available
  in the combinatorial case. In this paper, we propose and analyze UMM, an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and unbalanced rank aggregation (uBorda). Experimental results on
  black-box versions of LOP, QAP and PFSP show that UMM is able to match, and
  sometimes surpass, the solutions obtained by CEGO, a Bayesian optimization
  algorithm for combinatorial optimization, specially when the budget of
  solutions evaluated approaches 400. The computational cost of UMM is at least
  an order of magnitude smaller than CEGO, which makes it specially useful when
  both computation time and number of evaluations are of concern.\sloppy
  \textcolor{red}{16 pages max, deadline: 1 November 2020}
  \keywords{Combinatorial optimization \and Bayesian optimization \and
    Expensive black-box optimization \and Estimation of distribution
    algorithms}
\end{abstract}

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}\Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\Sigma \assign$ sampled $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}\Comment{Generate $\beta$ permutations}
    \State $\sigma\assign  \arg\max_{\pi'\in \Sigma}\arg\min_{ \pi \in S} d(\pi', \pi)   $ \Comment{Select the farthest from $S$}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}), which can be seen as replicating each permutation
$\sigma \in S$ a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}) and
the sampling of a new permutation from $MM(\sigma_0,\theta)$ (line~\ref{line:sample}). % in Algorithm~\ref{alg:umm}).

@@

A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}. The algorithm also implements the following strategy to avoid sampling the same points more than once, which will be more likely at the explotation phase. The strategy is as follows. Instead of sampling just one permutation from the MM, the algorithm generates $\beta$ permutations, $\Sigma$. Then it selects  for evaluation the farthest to the sample, i.e.,  $\sigma\in \Sigma$ such that $\sigma = \arg\max\arg\min d(\sigma, \pi)$ for every $\pi\in S$.

@@

\subsubsection{CEGO}
%
We use the implementation of GECO provided by the
authors.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of
$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
computation time, which was infeasible for our study. Due to extremely long
runtimes of CEGO and constraints in our computing system, we set the budget of
the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
Although a higher GA budget could improve the results of CEGO reported here,
our results are similar (and sometimes better) than the ones reported in the
original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. In fact, we found that
doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values, e.g., 0.5,
 show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}

>>>>

\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Black-box combinatorial optimization problems arise in practice when the
  objective function is evaluated by means of a simulator or a real-world
  experiment. In such cases, classical techniques such as mixed-integer
  programming and local search cannot be applied. Moreover, each fitness
  evaluation is often expensive in terms of time or resources, thus only a
  limited number of evaluations is possible, typically several orders of
  magnitude smaller than in white-box optimization problems. In the continuous
  case, Bayesian optimization, in particular using Gaussian processes, has
  proven very effective under these conditions. Much less research is available
  in the combinatorial case. In this paper, we propose and analyze UMM, an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and unbalanced rank aggregation (uBorda). Experimental results on
  black-box versions of LOP, QAP and PFSP show that UMM is able to match, and
  sometimes surpass, the solutions obtained by CEGO, a Bayesian optimization
  algorithm for combinatorial optimization, specially when the budget of
  solutions evaluated approaches 400. The computational cost of UMM is at least
  an order of magnitude smaller than CEGO, which makes it specially useful when
  both computation time and number of evaluations are of concern.\sloppy
  %\textcolor{red}{16 pages max, deadline: 1 November 2020}
  \keywords{Combinatorial optimization \and Bayesian optimization \and
    Expensive black-box optimization \and Estimation of distribution
    algorithms}
\end{abstract}

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}\Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
    \State $\Sigma \assign$ sample $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}\Comment{Generate $\beta$ permutations}
    \State $\sigma\assign  \argmax_{\pi'\in \Sigma}\min_{ \pi \in S} d(\pi', \pi)   $ \Comment{Select the farthest from $S$}\label{line:distance}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}), which can be seen as replicating each permutation
$\sigma \in S$ a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}) and
the sampling from $MM(\sigma_0,\theta)$ (lines~\ref{line:sample}--\ref{line:distance}). % in Algorithm~\ref{alg:umm}).

@@

A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.

To avoid sampling the same points more than once and increase the exploration of the search space, the algorithm samples $\beta$ permutations, $\Sigma$. Then it selects  for evaluation the farthest from permutations already evaluated ($S$), as shown in line~\ref{line:distance} of Algorithm~\ref{alg:umm}.
%i.e.,  $\sigma\in \Sigma$ such that $\sigma = \argmax_{\sigma'\in \Sigma}\min d(\sigma', \pi)$ for every $\pi\in S$.

@@

\subsubsection{CEGO}
%
We use the original implementation of GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of
$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
computation time, which was infeasible for our study. Due to extremely long
runtimes of CEGO and constraints in our computing system, we set the budget of
the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
Although a higher GA budget could improve the results of CEGO reported here,
our results are similar (and sometimes better) than the ones reported in the
original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. We found that
doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling of 10 solutions, with are quite poor as
expected, followed by building and updating the Kriging model based new
evaluations, which leads to an impressive improvement in less than 50
evaluations. However, progress is halted at this point and the model seems to
have trouble generating better solutions. Around evaluation 200, the search
becomes fully random. Lower uniformity values, e.g., 0.5, show worsening
evaluations despite the increase in solutions used to build the model. In
summary, the Kriging model clearly helps CEGO to quickly identify good
solutions in very few evaluations. However, for some reason, CEGO is not able
to further improve those solutions given more data. Given what we know about
Kriging~\citep{EriPeaGar2019scalable}, it appears that the surrogate model cannot model the local step
necessary to reach the optimal solution and switches to an exploration strategy
that ends up being completely random.

% This may due to the model not being able to produce accurate
% predictions or the underlying optimizer not being able to find improved
% solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}
